Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-18 tex/manuscript.tex
Appendiks fullført, ny figur, diff oppdatert

A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{q,n-q-1}$ distribution. However, if the data have been batch adjusted prior to this analysis, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
\begin{equation}
F\approx\frac{\tilde q\tilde\sigma^2}{q\sigma^2}\cdot\left(1+\frac{r}{n-q-r-1}\right)\cdot\text{F}_{\tilde q,n-q-r-1}
\end{equation}
where $\tilde q$ and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are perfectly balanced between the batches.

@@

A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made. A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size. In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups. A similar computation, although more complicated, can be made for multiple groups \NB{refer to appendix for details}. However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons. If the $\nu_0/\nu$ ratio is close to 1, results should still be reliable, while a ratio much larger than 1 will indicate exaggerated confidence. If the behaviour of the test statistic is well understood, as the $F$-statistic in ANOVA, one may adjust the statistic according to the $\nu_0/\nu$ ratio and recompute the $p$-value. However, this will require deep insight into the statistical behaviour of the model. Down-sampling to the effective samples size, i.e. to a $\nu/\nu_0$ portion of the samples, may also be done to see if results persist. However, none of these solutions are ideal.

@@

If there are no batches, i.e. $r=0$, we get the common linear model in which we get a decomposition $Y=\hat\beta_0B_0+\hat\epsilon_0$ with $\hat\gamma_0=Y_0B_0^t(B_0B_0^t)^{-1}$, where
$|\hat\epsilon|^2\sim\sigma^2\chi_{n-p-q}$ and $|(\hat\beta_0-\beta_0)C_0|^2\sim\sigma^2\chi_q$, yielding the common
$F$-statistic
\begin{equation}
F=\frac{|(\hat\beta_0-\beta_0)C_0|^2/q}{|\hat\epsilon_0|^2/(n-p-q)}\sim\text{F}_{q,n-p-q}.
\end{equation}
If we believe our data are  ``batch effect free'', this is the model we would use, and the distribution of the $F$-statistic we would assume.

We define the batch adjusted data $\tilde Y_0=Y_0-\hat\gamma_0C_0$ using the estimated batch effect $\hat\gamma_0$ from the full model. If we let $X_0=[B_0^t\mid C_0^t]^t$ be the full design matrix after elimination of covariates $A$, the parameter estimates are $[\hat\beta_0\mid\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$. The linear decomposition $Y_0=\hat\beta_0B_0+\hat\gamma_0C_0+\hat\epsilon_0$ has $|\hat\epsilon_0|^2\sim\sigma^2\chi^2_{n-p-q-r}$ and
$|(\hat\beta_0-\beta_0)B_0+(\hat\gamma_0-\gamma_0)C_0|^2\sim\sigma^2\chi^2_{q+r}$, and so the variance of the error term is the same in the batch adjusted case $\tilde Y_0=\hat\beta_0C_0+\hat\epsilon_0$. However, the sum of squares $|(\hat\beta_0-\beta_0)B_0|^2$ explained by $B$ is less easily expressed.

Solving $[\hat\beta_0\mid\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$ and using
\begin{equation}
(X_0X_0^t)^{-1}
\begin{pmatrix}B_0\\0\end{pmatrix}
=
\begin{pmatrix}P\\-(C_0C_0^t)^{-1}C_0B_0^tP\end{pmatrix} B_0
\end{equation}
with $P=(B_0\rho_{C_0}B_0^t)^{-1}$ and $\rho_{C_0}=I-C_0(C_0C_0^t)^{-1}C_0$
yields
\begin{equation}
(\hat\beta_0-\beta_0)B_0=\epsilon_0\rho_{C_0}B_0^tPB_0\sim\text{N}(0,\sigma^2L_0)
\quad\text{with}\quad
L_0=B_0^tPB_0\rho_{C_0}B_0^tPB_0
\end{equation}
since $\epsilon_0\sim\text{N}(0,\sigma^2I)$. Let $\lambda_1,\ldots,\lambda_q\ge1$ be the non-zero eigenvalues of $L_0$: this holds since $L_0$ is a symmetric rank $q$ matrix \NB{complete the proof of this}. Then,
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2=\sigma^2\sum_{i=1}^q\lambda_iU_i
\quad\text{where}\quad
U_1,\ldots,U_q\sim\chi^2_1.
\end{equation}
Approximating this so that the first two momenta match gives
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2\approx\tilde\sigma^2\chi^2_{\tilde q}
\quad\text{where}\quad
\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\sum_{i=1}^q\lambda_i=\text{tr}(L_0),\,
\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\sum_{i=1}^q\lambda_i^2=\text{tr}(L_0^2),
\end{equation}
from which it follows that $\tilde\sigma^2\ge\sigma^2$, $\tilde q\tilde\sigma^2\ge q\sigma^2$, and $\tilde q\le q$, where equality holds if and only if all $\lambda_i=1$, which happens when $B_0$ and $C_0$ are linearly independent, i.e. $B_0C_0^t=0$.

We would like to express $\tilde\sigma^2$ and $\tilde q$ terms of the original design matrices. The rotated design matrices were defined as $B_0=BU_0$, etc., using the rotation $U=[U_0\mid U']$ so that $AU_0=0$ while $AU'$ is non-singular. Hence, it follows that $V_0V_0^t=\rho_A=1-A^t(AA^t)^{-1}A$ which is the projection onto the $n-p$-hyperplane perpendicular to the $p$-hyperplane spanned by $A$. This allows us to express
\begin{equation}
L_0=U_0^tLU_0
\quad\text{where}\quad
L=
\end{equation}

@@

\begin{figure}[!p]
\centering\includegraphics[width=13cm]{Fig/boxplots.pdf}
%\centering\includegraphics[width=13cm]{Fig/boxplots.png}
\caption{We simulated expression of one gene from four study groups unevenly distributed in three batches containing 50+0+0+20, 0+50+0+20 and 0+0+50+20 samples from groups 1 to 4. This case, design and effects, was selected to illustrate the spurious effects that may arise from different batch adjustments. The Y-axis represent the expression values, while the X-axis is used to visually separate the batches. Circles, triangles, and crosses indicate values from each of the three batches, with colours indicating study groups. Correspondingly coloured boxes to the right of the measurements show group means with 95\% confidence intervals. These give some indication as to which group differences would be found significant.
a) The ``true'' values measured in a system without batch effects. Groups 1 and 2 have lower means than groups 3 and 4.
b) The same samples as in a) but measured in a system with batch effects. All the groups seem different.
c) The measurements in b) are adjusted with batch centering. Group 3 and 4 seem to differ while group 1 and 3 are more similar.
d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Group 1 and 2 seem to differ.
e) The least squares estimates of the group means from a two-way ANOVA have the same means as in d), but more appropriate confidence intervals.
}
\label{fig:boxplots}
\end{figure}

>>>>

A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{q,n-q-1}$ distribution. However, if the data have been batch adjusted prior to this analysis using two-way ANOVA, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
\begin{equation}
F\sim\frac{\sum_{i=1}^q\lambda_i\chi^2_{1}}{\chi^2_{n-q-r-1}}
\approx\frac{\tilde q\tilde\sigma^2}{q\sigma^2}\cdot\left(1+\frac{r}{n-q-r-1}\right)\cdot\text{F}_{\tilde q,n-q-r-1}
\end{equation}
where the eigenvalues $\lambda_i$, $\tilde q$, and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are evenly represented in all batches.

@@

A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made. A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size. In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups. A similar computation, although more complicated, can be made for multiple groups: see appendix for details covering general linear models. However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons. If the $\nu_0/\nu$ ratio is close to 1, results should still be reliable, while a ratio much larger than 1 will indicate exaggerated confidence. If the behaviour of the test statistic is well understood, as the $F$-statistic in ANOVA, one may adjust the statistic according to the $\nu_0/\nu$ ratio and recompute the $p$-value. However, this will require deep insight into the statistical behaviour of the model. Down-sampling to the effective samples size, i.e. to a $\nu/\nu_0$ portion of the samples, may also be done to see if results persist. However, none of these solutions are ideal.

@@

If there are no batches, i.e. $r=0$, we get the common linear model in which we get a decomposition $Y=\hat\beta_0B_0+\hat\epsilon_0$ with $\hat\gamma_0=Y_0B_0^t(B_0B_0^t)^{-1}$, where
$|\hat\epsilon|^2\sim\sigma^2\chi^2_{n-p-q}$ and $|(\hat\beta_0-\beta_0)C_0|^2\sim\sigma^2\chi^2_q$, yielding the common
$F$-statistic
\begin{equation}
F=\frac{|(\hat\beta_0-\beta_0)C_0|^2/q}{|\hat\epsilon_0|^2/(n-p-q)}\sim\text{F}_{q,n-p-q}.
\end{equation}
If we believe our data are  ``batch effect free'', and therefore do not include batch in the model, this is the distribution of the $F$-statistic we would assume.

We define the batch adjusted data $\tilde Y_0=Y_0-\hat\gamma_0C_0$ using the estimated batch effect $\hat\gamma_0$ from the full model. If we let $X_0=[B_0^t| C_0^t]^t$ be the full design matrix after elimination of covariates $A$, the parameter estimates are $[\hat\beta_0|\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$. The linear decomposition $Y_0=\hat\beta_0B_0+\hat\gamma_0C_0+\hat\epsilon_0$ has $|\hat\epsilon_0|^2\sim\sigma^2\chi^2_{n-p-q-r}$ and
$|(\hat\beta_0-\beta_0)B_0+(\hat\gamma_0-\gamma_0)C_0|^2\sim\sigma^2\chi^2_{q+r}$, and so the variance of the error term is the same in the batch adjusted case $\tilde Y_0=\hat\beta_0C_0+\hat\epsilon_0$. However, the sum of squares $|(\hat\beta_0-\beta_0)B_0|^2$ explained by $B$ is less easily expressed.

Solving $[\hat\beta_0|\hat\gamma_0]=Y_0X_0^t(X_0X_0^t)^{-1}$ and using
\begin{equation}
X_0^t(X_0X_0^t)^{-1} \begin{bmatrix}B_0\\0\end{bmatrix}
=[B_0^t|C_0^t] \begin{bmatrix}P\\-(C_0C_0^t)^{-1}C_0B_0^tP\end{bmatrix} B_0
=\rho_{C_0}B_0^tPB_0
\end{equation}
with $P=(B_0\rho_{C_0}B_0^t)^{-1}$, and $\rho_{W}=I-W^t(WW^t)^{-1}W$ for any $k\times n$ matrix with $k\le n$ defining the projection to the $n-k$-hyperplane perpendicular to the $k$-hyperplane spanned by $W$, yields
\begin{equation}
(\hat\beta_0-\beta_0)B_0=\epsilon_0\rho_{C_0}B_0^tPB_0\sim\text{N}(0,\sigma^2L)
\quad\text{where}\quad
L=B_0^tPB_0
\end{equation}
since $\epsilon_0\sim\text{N}(0,\sigma^2I)$. Let $\lambda_1,\ldots,\lambda_q\ge1$ be the non-zero eigenvalues of $L$: this holds since $L$ is a symmetric rank $q$ matrix,  has the same eigenvalues as $PB_0B_0^t$ the inverse of which, $(B_0B_0^t)^{-1}B_0\rho_{C}B_0^t$, has eigenvalues in $(0,1)$ (not 0 since $B_0$ and $C_0$ are linearly independent). Then,
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2=\sigma^2\sum_{i=1}^q\lambda_iU_i
\quad\text{where}\quad
U_1,\ldots,U_q\sim\chi^2_1.
\end{equation}
Approximating this so that the first two momenta match gives
\begin{equation}
|(\hat\beta_0-\beta_0)B_0|^2\approx\tilde\sigma^2\chi^2_{\tilde q}
\quad\text{where}\quad
\frac{\tilde q\tilde\sigma^2}{\sigma^2}=\sum_{i=1}^q\lambda_i=\text{tr}(L),\,
\frac{\tilde q\tilde\sigma^4}{\sigma^4}=\sum_{i=1}^q\lambda_i^2=\text{tr}(L^2),
\end{equation}
from which it follows that $\tilde\sigma^2\ge\sigma^2$, $\tilde q\tilde\sigma^2\ge q\sigma^2$, and $\tilde q\le q$, where equality holds if and only if all $\lambda_i=1$, which happens when $B_0$ and $C_0$ are linearly independent, i.e. $B_0C_0^t=0$.

We would like to express $\tilde\sigma^2$ and $\tilde q$ terms of the original design matrices. The rotated design matrices were defined as $B_0=BU_0$, etc., using the rotation $U=[U_0|U']$ so that $AU_0=0$ while $AU'$ is non-singular. Hence, it follows that $V_0V_0^t=\rho_A=1-A^t(AA^t)^{-1}A$. This allows us to express
\begin{equation}
L=U_0^tB^tPBU_0
\quad\text{where}\quad
P=(BU_0\rho_{C_0}U_0^tB^t)^{-1}=(B\rho_{A,C}B^t)^{-1}
\end{equation}
and $\rho_{A,C}=\rho_{[A^t|C^t]^t}$ is the projection to the $n-p-r$-hyperplane perpendicular to the $p+r$-hyperplane spanned by $A$ and $C$. We can then compute
\begin{equation}
\text{tr}(L)=\text{tr}(M),\,
\text{tr}(L^2)=\text{tr}(M^2)
\quad\text{where}\quad
M=BU_0U_0^tB^tP=B\rho_A B^t(B\rho_{A,C}B^t)^{-1}
\end{equation}
and where $M$ and $L$ have the same eigenvalues.
Notice that if $B$ is orthogonal to $A$, which can be achieved by replacing $B$ with $B-VA$ for some matrix $V$, then $B\rho_A=B$. If $C$ is also made orthogonal to $A$, then $\rho_{A,C}$ can be replaced by $\rho_C$.

@@

\begin{figure}[!p]
\centering\includegraphics[width=8cm]{Fig/boxplots.pdf}
%\centering\includegraphics[width=13cm]{Fig/boxplots_old.pdf}% For the old version
%\centering\includegraphics[width=13cm]{Fig/boxplots.png}
\caption{We simulated expression of one gene from four study groups unevenly distributed in three batches containing 50+0+0+20, 0+50+0+20 and 0+0+50+20 samples from groups 1 to 4. This case, design and effects, was selected to illustrate the spurious effects that may arise from different batch adjustments. The Y-axis represent the expression values, while the X-axis is used to visually separate the batches. Circles, triangles, and crosses indicate values from each of the three batches, with colours indicating study groups. Correspondingly coloured boxes to the right of the measurements show group means with 95\% confidence intervals. These give some indication as to which group differences would be found significant.
a) The ``true'' values measured in a system without batch effects. Groups 1 and 2 have lower means than groups 3 and 4.
b) The same samples as in a) but measured in a system with batch effects. All the groups seem different.
c) The measurements in b) are adjusted with batch centering. Group 3 and 4 seem to differ while group 1 and 3 are more similar.
d) The measurements in b) are adjusted with two-way ANOVA based batch centering (using \texttt{limma}). Group 1 and 2 seem to differ.
e) The least squares estimates of the group means from a two-way ANOVA have the same means as in d), but more appropriate confidence intervals.
}
\label{fig:boxplots}
\end{figure}
